Enzo: An Adaptive Mesh Refinement Code for Astrophysics

### 2013-04-28 numerical-starformation.tex
Edited star formation section.  This needs to be looked at by somebody else, please!

@@

While the physics discussed in other sections of this paper is crucial
to the study of cosmological structure formation, most observations of
galaxies in the infrared, optical, and ultraviolet wavelengths are of
stellar populations.  Furthermore, stars eject energy and metal-
enriched gas throughout their lives, drastically modifying their own
environments' cooling properties.  Broadly speaking, the formation of
galaxies and galaxy clusters cannot be completely modeled without
including the feedback of energy and metals from stellar populations.
In particular, it is thought that feedback is crucial for suppressing
the large numbers of dwarf-like galaxies that CDM theories predict
\citep{1991ApJ...381...14L,1991ApJ...379...52W}.  An early burst of
star formation could remove a large fraction of cold gas from such
systems \citep{1978MNRAS.183..341W,1991ApJ...367...45C}.  Also, the
unexpectedly low luminosities of small clusters and groups (relative
to the richest clusters) is commonly explained through feedback
\citep{1991ApJ...383..104K}.  Energy ejected during the formation of
the cluster ellipticals lowers the central density and hence the X-ray
luminosity of such clusters \citep{1997ApJ...484L..21C}.  For all of
these reasons, it is critical to include star formation and the
feedback of energy and metals into the interstellar and intergalactic
media in cosmological simulations.

Modeling star formation in the context of galaxy formation forces one
to confront two fundamental problems. The dynamical range required to
study galaxy formation and evolution is substantial, and attempting to
directly resolve the formation of individual stars in the context of a
Milky Way-type galaxy is computationally unfeasible.  Also, a strong
theoretical understanding of the details of star formation is
currently lacking.  Due to these challenges, models of star formation
in cosmological simulations are largely phenomenological, and the
``stars'' that are formed are in most cases meant to represent an
ensemble stellar population (although not all; see
Section~\ref{sec:starform_pop3}).  % BWO: I use 'phenomenological'
right above this, but could probably % go for 'parametric' or
'heuristic' if people object to my usage.

The \enzo\ code includes approximately ten models for star formation
and feedback.  Broadly speaking, these methods all work in similar
ways: At a specified time interval, all grid cells that are at the
highest local level of refinement (i.e., have no child cells) are
examined to see if they meet a set of criteria for star formation.
This may simply be a baryon density threshold, but may include more
complex tests, such as an examination of local cooling and dynamical
time scales, molecular hydrogen fraction, metallicity, and converging
gas flows.  If the star formation criteria are met, some mass of gas
is taken away from the cell in question and a ``star particle'' with
the same mass is placed in the center of that cell, with the same
momentum vector as the removed gas.  This star particle is then
allowed to inject mass, momentum, thermal energy, metals, and possibly
magnetic fields and/or cosmic ray populations into its local
environment.  In general, the particle is treated as an ensemble of
stars, with feedback properties occuring over time according to the
assumed initial mass function of the stellar population.

@@

If $m_{*}$ is greater than a minimum star mass $m_{\rm *min}$, a particle
is created and given several attributes: mass, a unique index number,
the time of formation $t_{\rm form}$, the local dynamical free-fall time
$t_{\rm dyn}$ and the metallicity fraction of the baryon gas in the cell
$f_{\rm Zb}$.  There is a user-defined minimum dynamical time
$T_{\rm dyn,min}$ which is observationally motivated and affects the
feedback rates (see below).  The particle is placed in the center of
the cell and given the same peculiar velocity as the gas in the cell,
and is then treated in the same manner as the dark matter particles.
An amount of baryon gas corresponding to the new particle's mass is
removed from the cell.

@@

The feedback of energy and metals into the baryon gas is similar to
the Kravtsov feedback, with some important differences.

@@

Algorithmically, the star formation events in Kravtsov's code are
assumed to occur once every global time step $\Delta t_0 \leq 10^7$.
In cells where star formation is determined to occur (i.e. $\rho_{\rm
  gas} \geq \rho_{\rm SF}$), star particles with a mass of $m_* =
\dot{\rho}_* V_{\rm cell} \Delta t_0$ (where $V_{\rm cell}$ is the
volume of the mesh cell) are assumed to form instantaneously in a
manner similar to that described in Section~\ref{sec:starform_cen}.
The \enzo\ implementation of this algorithm is similar, except that
instead of forming stars only at the root grid time step, we allow
stars to form at the time step of the highest level of resolution at
any particular point in space.  As can be seen from the equation for
$m_*$ above, this can result in very small stellar masses.  To avoid
memory and processor time issues related to having very large numbers
of star particles we impose a threshold mass $M_{\rm *,min}$ such that
a star particle only forms if $m_* \geq M_{\rm *,min}$.  An
appropriate choice of $M_{\rm *,min}$ does not significantly change
the star overall star formation history of a simulation, though it may
delay the onset of star formation in a given cell relative to a
simulation without a particle mass threshold.

@@

The methods described in the previous two sections are generally used
in simulations that have very poor resolution, $\Delta x \ga 1$~kpc.
At this physical scale, individual star forming regions are not
resolved, so all uncertainty about the behavior of molecular clouds is
folded into a density threshold for star formation.  In calculations
with much higher resolution, however ( $\la$~few pc), individual
molecular clouds can be resolved, thus rendering these approximations
irrelevant.  To that end, \citet{2012ApJ...749...36K} have implemented
a star formation algorithm that is specifically geared to
high-resolution cosmological simulations of galaxy formation, where
the formation of molecular hydrogen is followed directly, and stars
are allowed to form at the highest level of refinement when the local
H$_2$ fraction exceeds a pre-determined threshold.  Cells are examined
every root grid timestep, $\Delta t_0$, and cells at the highest level
that exceed the H$_2$ fraction limit form star particles with masses
proportional to the inferred mass of molecular hydrogen in the
star-forming region
\citep{2008ApJ...689..865K,2009ApJ...693..216K,2010ApJ...709..308M}.
The mass of the particle is calculated as:

@@

Unlike in the previous sections, under some circumstances it is both
possible and desirable to simulate stars individually, rather than
treating particles as ensembles.  One particular example of this is
Population III star formation
\citep{ABN02,2007ApJ...654...66O,2008ApJ...685...40W,2009Sci...325..601T},
where a given halo may only form one star of primordial composition.
To accommodate this, Enzo contains a star formation algorithm that
forms Population III stars directly \citep{2007ApJ...659L..87A,
  2008ApJ...685...40W, 2012MNRAS.427..311W}.  Using criteria similar
to~\citet{CO1992}, a star particle forms when a cell has all of the
following criteria:

\begin{enumerate}
\item a baryon overdensity of $5 \times 10^5$ (corresponding to
  roughly $10^3$ g cm$^{-3}$ at $z=10$),

\item a converging gas flow ($\nabla \cdot \vec{v} < 0$), and

\item a molecular hydrogen fraction f$_{\rm H2} > 5 \times 10^{-4}$,
\end{enumerate}

conditions typical of a collapsing metal-free molecular cloud roughly
10 million years before the birth of a Pop III main-sequence star.  If
multiple neighboring cells are flagged as being able to form stars, a
single star is created instead.  This star has a mass that is randomly
sampled from a stellar IMF with a functional form of

@@

Feedback from the Population III stars comes in multiple forms.
Radiative feedback using the MORAY radiation transport algorithm
\citep{Wise11_Moray} is available, using the mass-dependent hydrogen
ionizing and Lyman-Werner photon luminosities and lifetimes of the
Population III stars from \citet{2002A&A...382...28S}.  At the end of
their main-sequence lifetimes, explosion energies, ejected gas mass,
and ejected metal are caclulated using a variety of sources that
depend on the mass of the star, and are described in detail in Section
3.2.1 of \citet{2012MNRAS.427..311W}.

>>>>

\red{( (BWO) I am a little worried that this section is too
  long. The editorial hatchet may be called for, but I think somebody
  else (who didn't write the section in the first place) needs to
  think it through. )}

@@

\red{(BWO: can this whole paragraph go?)}
While the physics discussed in other sections of this paper is crucial
to the study of cosmological structure formation, most observations of
galaxies in the infrared, optical, and ultraviolet wavelengths are of
stellar populations.  Furthermore, stars eject energy and metal-
enriched gas throughout their lives, drastically modifying their own
environments' cooling properties.  Broadly speaking, the formation of
galaxies and galaxy clusters cannot be completely modeled without
including the feedback of energy and metals from stellar populations.
In particular, it is thought that feedback is crucial for suppressing
the large numbers of dwarf-like galaxies that CDM theories predict
\citep{1991ApJ...381...14L,1991ApJ...379...52W}.  An early burst of
star formation could remove a large fraction of cold gas from such
systems \citep{1978MNRAS.183..341W,1991ApJ...367...45C}.  Also, the
unexpectedly low luminosities of small clusters and groups (relative
to the richest clusters) is commonly explained through feedback
\citep{1991ApJ...383..104K}.  Energy ejected during the formation of
the cluster ellipticals lowers the central density and hence the X-ray
luminosity of such clusters \citep{1997ApJ...484L..21C}.  For all of
these reasons, it is critical to include star formation and the
feedback of energy and metals into the interstellar and intergalactic
media in cosmological simulations.

Modeling star formation in the context of galaxy formation forces one
to confront two fundamental problems. The dynamical range required to
study galaxy formation and evolution is substantial, and attempting to
directly resolve the formation of individual stars in the context of a
Milky Way-type galaxy is computationally unfeasible.  Also, a strong
theoretical understanding of the details of star formation is
currently lacking.  Due to these challenges, models of star formation
in cosmological simulations are largely phenomenological, and the
``stars'' that are formed are in most cases meant to represent an
ensemble stellar population (although not all; see
Section~\ref{sec:starform_pop3}).

The \enzo\ code includes approximately ten models for star formation
and feedback.  Broadly speaking, these methods all work in similar
ways: At a specified time interval, all grid cells that are at the
highest local level of refinement (i.e., that have no child cells) are
examined to see if they meet a set of criteria for star formation.
This may simply be a baryon density threshold, but may include more
complex tests, such as an examination of local cooling and dynamical
time scales, molecular hydrogen fraction, metallicity, and converging
gas flows.  If the star formation criteria are met, some mass of gas
is taken away from the cell in question and a ``star particle'' with
the same mass is placed in the center of that cell, with the same
momentum vector as the removed gas.  This star particle is then
allowed to inject mass, momentum, thermal energy, metals, and possibly
magnetic fields and/or cosmic ray populations into its local
environment.  In general, the particle is treated as an ensemble of
stars, with feedback properties occuring over time according to the
assumed initial mass function of the stellar population.

@@

If $m_{*}$ is greater than a minimum star mass $m_{\rm *min}$, a particle
is created and given several attributes: mass, a unique index number,
the time of formation $t_{\rm form}$, the local dynamical free-fall time
$t_{\rm dyn}$ and the metallicity fraction of the baryon gas in the cell
$f_{\rm Zb}$.  There is a user-defined minimum dynamical time
$T_{\rm dyn,min}$, which is observationally motivated and affects the
feedback rates (see below).  The particle is placed in the center of
the cell and given the same peculiar velocity as the gas in the cell,
and is then treated in the same manner as the dark matter particles.
An amount of gas corresponding to the new particle's mass is
removed from the cell.

@@

% The feedback of energy and metals into the baryon gas is similar to
% the Kravtsov feedback, with some important differences.

@@

Algorithmically, the star formation events in Kravtsov's code are
assumed to occur once every global time step (with the constraint $\Delta t_0 \leq
10^7$ years).
In cells where star formation is determined to occur (i.e. $\rho_{\rm
  gas} \geq \rho_{\rm SF}$), star particles with a mass of $m_* =
\dot{\rho}_* V_{\rm cell} \Delta t_0$ (where $V_{\rm cell}$ is the
volume of the mesh cell) are assumed to form instantaneously in a
manner similar to that described in Section~\ref{sec:starform_cen}.
The \enzo\ implementation of this algorithm is similar, except that
instead of forming stars only at the root grid time step, we allow
stars to form at the time step of the highest level of resolution at
any particular point in space.  As can be seen from the equation for
$m_*$ above, this can result in very small stellar masses.  To avoid
memory and processor time issues related to having very large numbers
of star particles we impose a threshold mass $M_{\rm *,min}$ such that
a star particle only forms if $m_* \geq M_{\rm *,min}$.  An
appropriate choice of $M_{\rm *,min}$ does not significantly change
the star overall star formation history of a simulation, though it may
delay the onset of star formation in a given cell relative to a
simulation without a particle mass threshold.

@@

The methods described in the previous two sections are generally used
in simulations that have very poor resolution, $\Delta x \ga 1$~kpc.
At this physical scale, individual star forming regions are not
resolved, so all uncertainty about the behavior of molecular clouds is
folded into a density threshold for star formation.  In calculations
with much higher resolution, however (on the order of a few pc), individual
molecular clouds can be resolved, thus rendering these approximations
invalid.  To that end, \citet{2012ApJ...749...36K} have implemented
a star formation algorithm that is specifically geared to
high-resolution cosmological simulations of galaxy formation, where
the formation of molecular hydrogen is followed directly, and stars
are allowed to form at the highest level of refinement when the local
H$_2$ fraction exceeds a pre-determined threshold.  Cells are examined
every root grid timestep, $\Delta t_0$, and cells at the highest level
that exceed the H$_2$ fraction limit form star particles with masses
proportional to the inferred mass of molecular hydrogen in the
star-forming region
\citep{2008ApJ...689..865K,2009ApJ...693..216K,2010ApJ...709..308M}.
The mass of the particle is calculated as:

@@

Unlike in the previous sections, under some circumstances it is both
possible and desirable to simulate stars individually, rather than
treating particles as ensembles.  One particular example of this is
Population III star formation
\citep{ABN02,2007ApJ...654...66O,2008ApJ...685...40W,2009Sci...325..601T},
where a given halo may only form one star of primordial composition.
To accommodate this, \enzo\ contains a star formation algorithm that
forms individual Population III stars directly \citep{2007ApJ...659L..87A,
  2008ApJ...685...40W, 2012MNRAS.427..311W}.  Using criteria similar
to~\citet{CO1992}, a star particle forms when a cell meets all of the
following conditions:

\begin{enumerate}
\item A baryon overdensity of $5 \times 10^5$ (corresponding to a
  hydrogen number density of 
  roughly $10^3$ cm$^{-3}$ at $z=10$),

\item A converging gas flow ($\nabla \cdot \vec{v} < 0$), and

\item A molecular hydrogen mass fraction f$_{\rm H2} > 5 \times 10^{-4}$,
\end{enumerate}

These are comparable to the conditions typical of a collapsing metal-free molecular cloud roughly
10 million years before the birth of a Pop III main-sequence star.  If
multiple neighboring cells are flagged as being able to form stars, a
single star is created instead.  This star has a mass that is randomly
sampled from a stellar IMF with a functional form of

@@

Feedback from Population III stars created using this algorithm comes in multiple forms.
Radiative feedback using the \moray\ radiation transport algorithm
\citep{Wise11_Moray} is available, using the mass-dependent hydrogen
ionizing and Lyman-Werner photon luminosities and lifetimes of the
Population III stars from \citet{2002A&A...382...28S}.  At the end of
their main-sequence lifetimes, explosion energies, ejected gas mass,
and ejected metal are caclulated using a variety of sources that
depend on the mass of the star, and are described in detail in Section
3.2.1 of \citet{2012MNRAS.427..311W}.
